---
title: "Compare exposure risk with prevalence rate EDA"
author: "Emma Pendl-Robinson"
date: "1/4/2021"
output: 
  html_document:
    toc: true
    number_sections: true 
    toc_float: 
      collapsed: false
      smooth_scroll: false
    css: style.css
    code_folding: hide 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Compare exposure risk with prevalence rate {-}


```{r libraries, warning=FALSE, message=FALSE}
library(tidyverse)
library(tidycensus)
library(magrittr)
```

```{r}
# read in covid_df data
covid_df <- read_rds(here::here("exposure-map/data/covid_transformed.rds"))

# read in fips codes
fips_state_xwalk <- read_csv(here::here("exposure-map/data/usafacts.csv")) %>%
  distinct(fips, state)

# add fips code
covid_df %<>% left_join(fips_state_xwalk, by = "fips")
```

# EDA

Finding similarities and and differences counties

```{r}
# add cases break points 
covid_df$cases_nytbreaks  <- cut(covid_df$cases_daily_100k, breaks = c(0, 10, 30, 50, 70, 100, 250, 700), include.lowest = TRUE)
covid_df$cases_cat <- as.numeric(covid_df$cases_nytbreaks)
# add exposure break points 
covid_df$expo_nytbreaks <-  cut(covid_df$exposure_risk_pct, breaks = c(0, 0.6, 1.2, 1.8, 2.4, 3, 4, 25), include.lowest = TRUE)
covid_df$expo_cat <- as.numeric(covid_df$expo_nytbreaks)

# check if similar
covid_df$cases_expo_cat <- covid_df$expo_cat - covid_df$cases_cat
```


# Example County Data


-	Alameda county, CA (fips 06001) 

```{r}
covid_df %>% filter(fips == "06001")

```
-	Alleghany County, VA (fips 51005)

```{r}
covid_df %>% filter(fips == "51005")
```

# Inter-state Varriance: spearman correlation
spearman correlation of the mean pervalance and exposure between states
```{r}
states_means <- covid_df %>%
  group_by(state) %>%
  summarise(mean_cases = mean(cases_daily_100k),
            mean_expo = mean(exposure_risk_pct),
            mean_cases_cat = mean(cases_cat),
            mean_expo_cat = mean(expo_cat),
            .groups = "drop")  

# comparing mean bins
cor.test(states_means$mean_cases_cat, states_means$mean_expo_cat, method = "spearman", exact = FALSE)

# comparing mean number
cor.test(states_means$mean_cases, states_means$mean_expo, method = "spearman", exact = FALSE)

```

# Intra-State 

This seems a bit repetitive of the Utah vs Texas example. Are you trying to say something about the intra-state variability here? Can you add to this by citing quantitative numbers? For example, county A and county B in North Carolina have similar case count per 100k population, XXX and XXX. However, their underreporting factor vary, and therefore the estimated exposure risk in county A and county B are XXX and XXX.

```{r}
nc_df <- covid_df %>% 
  filter(state == "NC") %>%
  select(fips, state, underreport_fac, cases_daily_100k, exposure_risk_pct, cases_cat, expo_cat) %>%
  mutate(mean_cases = mean(cases_daily_100k),
         mean_expo = mean(exposure_risk_pct)) 
```


Foundes with about average cases in North Carolina
Bertie County, NC (fips 37015)

Mecklenburg County, NC (fips 37119)
```{r}
nc_df %>% filter(fips %in% c("37015", "37119"))

```




