---
title: "Compare exposure risk with prevalence rate"
author: "Emma Pendl-Robinson"
date: "1/4/2021"
output: 
  html_document:
    toc: true
    number_sections: true 
    toc_float: 
      collapsed: false
      smooth_scroll: false
    css: style.css
    code_folding: hide 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Compare exposure risk with prevalence rate {-}

Exposure risk and how it changes overtime

Looking at prevalence and make a map

See if the exposure risk we should in different places vs the exposure map

Doing this in map form


```{r libraries, warning=FALSE, message=FALSE}
source(here::here("exposure-map/functions_n_packages.R"))
library(magrittr)
```



# Data Processing

## make data set:

```{r, eval = FALSE}
usafacts_df <- get_data()

write_csv(usafacts_df, here::here("exposure-map/data/usafacts.csv"))
```


read in USA facts data

```{r}
usafacts_df <- read_csv(here::here("exposure-map/data/usafacts.csv"))

# filter to date less than 2020/12/15
usafacts_df %<>% filter(date <= ymd("2020-12-15"))

```

## Calc CFR 

Use the EB adjusted state CFR priors

```{r}
# Priors downloaded from  https://github.com/mathematica-mpr/math-api-covid-risk-calculator-etl/blob/7574f08ccac1c345b1ce54a2f2f15f4306734773/risk_score_etl/risk_score_etl/resources/state_CFR_priors.csv 

state_CFR_priors <- read.csv(here::here("exposure-map/data/state_CFR_priors.csv"))

```

Adjust CFR

TODO: 

    - fix fips with negativee deaths and CFR values
        - for 6 fips the total number of deaths was higher 90 days ago and have a negative number of deaths
        - for now, I will set these CFRs and deaths to zero
    - what priors should I use? 
        - Erin just redid the priors
```{r}

update_prior_shape_df <- shape_data(usafacts_df) %>%
  mutate(CFR = ifelse(CFR < 0, 0 , CFR),
         deaths = ifelse(deaths < 0, 0 , deaths))

update_prior_shape_df$CFR_adj <- calc_adjusted_CFR(update_prior_shape_df)

fips_priors <- left_join(update_prior_shape_df, state_CFR_priors, by = "state") %>%
    mutate(CFR_adj_EB = (deaths+shape1)/(cases+shape1+shape2),
           case_cat = case_when(cases>=1000~"1000+",
                                cases>=100~"100-999",
                                cases<100~"0-99"),
           CFR = ifelse(cases==0, 0, CFR)) %>%
    filter(state!="DC")

```

Calc Underreporting Factor

- Use the adjusted CFR from above to caculate the underreporting factor when IFR is set to 0.006).

```{r}
# deaths is deaths in the last 90 days
# cases from day 90 last 90 days with 13 day lag
covid_underreporting_df <- fips_priors %>% 
  mutate(underreport_fac = CFR_adj_EB/0.006) %>%
  mutate(underreport_fac = case_when(underreport_fac < 1 ~ 1,
                                     underreport_fac > 10 ~ 1,
                                     TRUE ~ underreport_fac)) %>%
  dplyr::select(fips, deaths, cases, underreport_fac) %>%
  rename("deaths_last90" = "deaths", "cases_last90" = "cases")

```

## current Cases

Find number of: 

- reported cases in last 14 days 
- reported deaths in last 14 days
- reported cases in last 7 days (1 week)

```{r}
covid_recent_days <- usafacts_df %>% 
  gather(var,count,c("cases","deaths")) %>%
  mutate(d_label = case_when(var=="deaths" & date == max(date) ~ "deaths_today",
                             var=="deaths" & date == max(date)-14 ~ "deaths_14",
                             var=="cases" & date == max(date) ~ "cases_today",
                             var=="cases" & date == max(date)-14 ~ "cases_14",
                             var=="cases" & date == max(date)-07 ~ "cases_07")) %>%
  filter(!is.na(d_label)) %>% dplyr::select(fips,state,count,d_label) %>%
  spread(d_label, count) %>%
  mutate(cases_last14 = cases_today - cases_14, 
         deaths_last14 = deaths_today-deaths_14,
         cases_last07 = cases_today - cases_07) %>%
  mutate(cases_last14 = ifelse(cases_last14 < 0, 0 , cases_last14),
         deaths_last14 = ifelse(deaths_last14 < 0, 0 , deaths_last14),
         cases_last07 = ifelse(cases_last07 < 0, 0 , cases_last07)) %>%
  dplyr::select(fips, cases_last14, deaths_last14, cases_last07)

```

calculate:

- `active casecount` (number of covid infections in the last 14 days) by the number cases in the last 14 days and the underreporting factor

```{r}
covid_df <- left_join(covid_underreporting_df, covid_recent_days, by = "fips") %>%
  mutate(active_casecount = underreport_fac*cases_last14)
  
```

calculate prevalance of covid infections in each county (fips code area)
```{r}
library(tidycensus)

census_api_key(Sys.getenv("census_api_key"))
v19 <- load_variables(2019, "acs5", cache = TRUE)

# find population in the fips area
fips_pop <- tidycensus::get_acs(geography = "county", variables = "B01003_001", year = 2019) %>%
  rename("population" = "estimate",
         "fips" = "GEOID") %>%
  dplyr::select(fips, population)

# calc infection and deaths per person people in the county
covid_df %<>%
  left_join(fips_pop, by = "fips") %>%
  mutate(prev_active = active_casecount/population)
```

caclulate risk of contracting COVID thorugh community if someone comes in contact with 10 people as percent

```
https://github.com/mathematica-mpr/covid_risk_score/blob/62ebd0a51dd126df8068a3d52a058881a7eeb45a/src/results.R 
exposure_risk <- 1-(1-prev_active*transmissibility_household)^(input$nppl+input$nppl2*transmissibility_household)

```

```{r}
# TODO : why uyse transmissibility_household vs the standard transmissibility
transmissibility_household <-0.105 

covid_df %<>%
  mutate(exposure_risk = 1-(1-prev_active*transmissibility_household)^(10)) %>%
  mutate(exposure_risk_pct = exposure_risk*100)
```

Calculate
`cases_daily_100k` Average daily cases per 100,000 people in past week

```{r}
covid_df %<>%
  mutate(cases_av_daily = cases_last07/7,
         cases_daily_100k = (cases_av_daily/population)*100000)
```


save data
```{r}
write_rds(covid_df, here::here("exposure/data/covid_transformed.rds"))
```

