---
title: "Compare exposure risk with prevalence rate"
author: "Emma Pendl-Robinson"
date: "1/4/2021"
output: 
  html_document:
    toc: true
    number_sections: true 
    toc_float: 
      collapsed: false
      smooth_scroll: false
    css: style.css
    code_folding: hide 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Compare exposure risk with prevalence rate {-}

Exposure risk and how it changes overtime

Looking at prevalence and make a map

See if the exposure risk we should in different places vs the exposure map

Doing this in map form


```{r libraries, warning=FALSE, message=FALSE}
library(tidyverse)
library(tidycensus)
library(magrittr)
```

```{r}
covid_df <- read_rds(here::here("exposure/data/covid_transformed.rds"))
```


# map data


## urban instutute mapping

https://urban-institute.medium.com/how-to-create-state-and-county-maps-easily-in-r-577d29300bb2

```{r}
library(urbnmapr)

# lat long codinates of counties in the uS
county_geo <- get_urbn_map("counties")

```

```{r}
# colors used by the NYTimes average daily cases per 100,000 people in the past week 
nyt_cols <- c("#f2df91", "#f9c467", "#ffa83e", "#ff8b24", "#fd6a0b", "#f04f09", "#d8382e", "#c62833", "#AF1C43", "#8A1739", "#701547", "#4C0D3E")

covid_map_style <- function(gplot){
  # styles ggplot
  gplot + 
  geom_polygon(color = NA) +
  geom_polygon(data = urbnmapr::states, mapping = aes(long, lat, group = group),
               fill = NA, color = "#ffffff") +
  coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
  scale_fill_manual(values = nyt_cols[c(1, seq(0, 12, 2))]) + 
  theme(plot.title = element_text(hjust = 0.5,
                                  size = 20),
        axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        panel.background = element_rect(fill = NA),
        legend.position= "top",
        legend.title = element_text(size = 10), 
        legend.key.size = unit(.8,"line"),
        legend.text = element_text(family = "mono",
                                   face = "bold",
                                   size = 7) ,
        plot.background = element_rect(color = "black")) + 
  guides(fill = guide_legend(nrow = 1,
                             title.position = "top", 
                             title.hjust = 0.5,
                             label.position = "bottom"))
}
```

## Daily Active cases

- `cases_dailyav` Average daily cases per 100,000 people in past week

NYTimes shows average daily cases per 100,000 people in the past week with intervals of: 10, 30, 50, 70, 100, and 250

- https://www.nytimes.com/interactive/2020/us/coronavirus-us-cases.html


```{r}
# breakpoints for interval
covid_df$cases_nytbreaks  <- cut(covid_df$cases_daily_100k, breaks = c(0, 10, 30, 50, 70, 100, 250, 700), include.lowest = TRUE)

# string pad to make lables even length and relevel http://www.cookbook-r.com/Manipulating_data/Renaming_levels_of_a_factor/
levels(covid_df$cases_nytbreaks) <- str_pad(levels(covid_df$cases_nytbreaks), 9, side = "both")

cases_plot <- left_join(covid_df, county_geo, by = c("fips" = "county_fips"))  %>%
  ggplot(aes(long, lat, group = group, fill = cases_nytbreaks)) +
  labs(title = "\nA) Reported Cases",
       fill = "Average daily cases per 100,000 people in past week")

cases_plot_styled <- covid_map_style(cases_plot)


```

## Exposure Risk

map of exposure risk (active cases/population)

```{r}
# breakpoints for interval
covid_df$expo_nytbreaks <-  cut(covid_df$exposure_risk_pct, breaks = c(0, 0.6, 1.2, 1.8, 2.4, 3, 4, 25), include.lowest = TRUE)

# string pad to make lables even length and relevel http://www.cookbook-r.com/Manipulating_data/Renaming_levels_of_a_factor/
levels(covid_df$expo_nytbreaks) <- str_pad(levels(covid_df$expo_nytbreaks), 9, side = "both")

exposure_plot <- left_join(covid_df, county_geo, by = c("fips" = "county_fips"))  %>%
  ggplot(aes(long, lat, group = group, fill = expo_nytbreaks)) +
  labs(title = "\nB) Exposure Risk",
       fill = "Exposure risk (%) for 10 direct contacts in next week")


exposure_plot_styled <- covid_map_style(exposure_plot)


```


## combine plots


http://www.sthda.com/english/articles/32-r-graphics-essentials/126-combine-multiple-ggplots-in-one-graph/

```{r}
# combine plots
combined_map <- ggpubr::ggarrange(cases_plot_styled, exposure_plot_styled,
                    labels = c("", ""),
                    ncol = 1, nrow = 2)

figures_filer <- "N:/Project/11123_19andMe/CA1/covid_risk_score_paper/figures/_paper-val-fig4.png"
#save plot
ggsave(combined_map,
       filename = figures_filer,
       width = 7, height = 10)

```



