---
title: "Validation of 19andMe against COVER"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, cache=F)
library(tidyverse)
library(pander)
library(ggthemes)

dat <- read_csv(file.path("..", "output-cover", "res-cover.csv"))

alg <- read_csv(file.path("..","input-cover","cover_algorithm.csv"))
COND_VARS_COVER <- pull(na.omit(filter(alg, risk_factor=="conditions"), "var"), "var")

COND_VARS_ALL <- c("is_renal", "is_cvd", "is_diabetes", "is_hyper", "is_immune",  
                   "is_lung", "is_smoker", "is_other", "is_obesity")
```

```{r}
# process results
COVER_VARS <- c("COVER_H","COVER_I","COVER_F")
APP_VARS <- c("api_hosp_risk", "api_icu_risk", "api_death_risk")

dat <- dat %>%
  rowwise() %>% 
  mutate(conditions_num = sum(is_renal, is_cvd, is_diabetes, is_hyper, is_immune, 
                              is_lung, is_smoker, is_other, is_obesity)) %>%
  ungroup()

res <- dat %>% 
  select(c(COVER_VARS, APP_VARS)) %>%
  set_names( c(paste0(c("hosp","icu","death"), "_cover"),
               paste0(c("hosp","icu","death"), "_MPR") )) %>%
  mutate(has_condition = !is.na(dat$conditions)) %>%
  bind_cols(select(dat, conditions_num, age, sex, COND_VARS_ALL))

res_long <- res %>%
  mutate(rownum = row_number()) %>%
  gather("key","score",c(-rownum, -has_condition, -conditions_num, 
                         -age, -sex, -is_other, -COND_VARS_ALL)) %>%
  separate(key, c("outcome", "app"), "_") %>%
  spread(app, score) %>%
  mutate(outcome = fct_relevel(outcome, c("hosp","icu","death")),
         outcome = case_when(outcome=="hosp"~"Hospitalization",
                             outcome=="icu"~"ICU",
                             outcome=="death"~"Mortality"),
         diff_cat = factor(case_when(MPR-cover>.1 ~ '19andMe 10% higher',   
                                     cover-MPR>.1 ~ '19andMe 10% lower',   
                                     TRUE ~ "within 10%"),
                              levels=c('within 10%', '19andMe 10% lower', '19andMe 10% higher'))) %>%
  group_by(outcome) %>%
  mutate(scorr = round(cor(MPR, cover, method="spearman"),2))
```

## Summarise validation dataset (subset)

### Age and sex
```{r}
ggplot(dat, aes(x=as.character(age), fill=sex)) +
  geom_bar(position = "dodge") +
  theme_bw()
```

### Count of conditions
```{r}
dat %>%
  ggplot(aes(x=conditions_num)) +
  geom_bar() +
  xlab("Number of conditions for COVER app") +
  theme_bw()
```

### Individual condition counts
```{r}
select(dat, COND_VARS_ALL) %>%
  gather("condition", ind) %>%
  filter(ind==1) %>%
  
  ggplot(aes(x=condition)) +
  geom_bar() +
  theme_bw()
```


## Compare COVER and 19andMe suseptability scores

### Spearman correlations
```{r}
res %>% select(matches("MPR|cover")) %>%
  cor(method="spearman") %>%
  as.data.frame() %>%
  rownames_to_column("var1") %>%
  gather("var2","cor",-var1) %>%
  mutate_at(c("var1", "var2"), fct_relevel, levels = names(res)) %>%
  
  ggplot(aes(x=var1, y=var2, fill=cor)) +
  geom_tile() +
  geom_text(aes(label = round(cor, 2))) +
  geom_vline(xintercept = 3.5) +
  geom_hline(yintercept = 3.5) +
  xlab("") + ylab("") +
  theme_bw()
```

### Correlation with p values
```{r, results="asis}
hosp_corrp <- cor.test(res$hosp_cover, res$hosp_MPR, method="spearman") %>%
  .$p.value %>% format.pval()
icu_corrp <- cor.test(res$icu_cover, res$icu_MPR, method="spearman") %>%
  .$p.value %>% format.pval()
death_corrp <- cor.test(res$death_cover, res$death_MPR, method="spearman") %>%
  .$p.value %>% format.pval()

tibble(outcome = c("hosp","icu","death"),
       corr_pval = c(hosp_corrp, icu_corrp, death_corrp)) %>% pandoc.table()
```

### Scatterplots
```{r}
res_long %>%
  ggplot(aes(y=MPR, x=cover)) +
  geom_point(size=.5) +
  geom_abline() +
  geom_abline(intercept = -.1, linetype="dotted") +
  geom_abline(intercept = .1, linetype="dotted") +
  
  facet_grid(.~outcome) +
  coord_fixed() +
  
  scale_y_continuous(lim=c(0,1), "19andMe probability", labels = scales::percent) +
  scale_x_continuous(lim=c(0,1), "COVER probability", labels = scales::percent) +
  theme_bw()
```

### Scatterplots - no conditions
```{r}
filter(res_long, !has_condition) %>%
  ggplot(aes(y=MPR, x=cover)) +
  geom_point(size=.5) +
  
  geom_abline() +
  geom_abline(intercept = -.1, linetype="dotted") +
  geom_abline(intercept = .1, linetype="dotted") +
  
  facet_grid(.~outcome) +
  coord_equal() +
  
  scale_y_continuous(lim=c(0,.3), "19andMe probability", labels = scales::percent) +
  scale_x_continuous(lim=c(0,.3), "COVER probability", labels = scales::percent) +
  theme_bw()
```

### Base score by age and sex
```{r, fig.width=8, fig.height=3}
filter(res_long, !has_condition) %>%
  gather("app","score", c(MPR, cover)) %>%
  filter(app!="MPR" | age%%10==0 & age<=80) %>%
  
  ggplot(aes(x=age, y=score, color=app, shape=sex)) +
  geom_point() +
  
  facet_grid(.~outcome) +
  scale_y_continuous(labels = scales::percent) +
  scale_color_discrete(labels = c("COVER","19andMe")) +
  theme_bw()
```

### Percent within each difference category
```{r, results="asis"}
res_long %>%
  count(outcome, diff_cat) %>%
  group_by(outcome) %>%
  mutate(p=n/sum(n)) %>% select(-n) %>%
  spread(diff_cat, p) %>%
  select(c(1,3,2,4)) %>%
  mutate_at(2:4, ~ifelse(is.na(.), 0, .)) %>%
  pandoc.table()
```

### Number within each difference category
```{r, results="asis"}
res_long %>%
  count(outcome, diff_cat) %>%
  spread(diff_cat, n) %>%
  select(c(1,3,2,4)) %>%
  mutate_at(2:4, ~ifelse(is.na(.), 0, .)) %>%
  pandoc.table()
```

### Percent within each difference category by age group
```{r, results="asis"}
res_long %>%
  mutate(age_cat = factor(case_when(age>=80 ~ "80+",
                             age>=60 ~ "60-79",
                             TRUE ~ "Under 60"),
                          levels = c("Under 60","60-79","80+")),
         age_cat_bin = if_else(age>=60, "60+", "Under 60")) %>%
  count(outcome, age_cat_bin, diff_cat) %>%
  group_by(outcome, age_cat_bin) %>%
  mutate(p=scales::percent(n/sum(n))) %>% select(-n) %>%
  spread(diff_cat, p) %>%
  select(c(1,2,4,3,5)) %>%
  mutate_at(3:5, ~ifelse(is.na(.), scales::percent(0), .)) %>%
  pandoc.table(split.table=Inf)
```

### Percent within each difference category by number of conditions
```{r, results="asis"}
res_long %>%
  mutate(cond_cat = case_when(conditions_num==0 ~ "0 conditions",
                              conditions_num==1 ~ "1 condition",
                              conditions_num==2 ~ "2 conditions",
                              TRUE ~ "3+ conditions")) %>%
  count(outcome, cond_cat, diff_cat) %>%
  group_by(outcome, cond_cat) %>%
  mutate(p=scales::percent(n/sum(n))) %>% select(-n) %>%
  spread(diff_cat, p) %>%
  select(c(1,2,4,3,5)) %>%
  mutate_at(3:5, ~ifelse(is.na(.), scales::percent(0), .)) %>%
  pandoc.table(split.table=Inf)
```

### Percent within each difference category by number of conditions - 60+ only
```{r, results="asis"}
res_long %>%
  filter(age>=60) %>%
  mutate(cond_cat = case_when(conditions_num==0 ~ "0 conditions",
                              conditions_num==1 ~ "1 condition",
                              conditions_num==2 ~ "2 conditions",
                              TRUE ~ "3+ conditions")) %>%
  count(outcome, cond_cat, diff_cat) %>%
  group_by(outcome, cond_cat) %>%
  mutate(p=scales::percent(n/sum(n))) %>% select(-n) %>%
  spread(diff_cat, p) %>%
  select(c(1,2,4,3,5)) %>%
  mutate_at(3:5, ~ifelse(is.na(.), scales::percent(0), .)) %>%
  pandoc.table(split.table=Inf)
```


## For paper

### Plot 19andMe versus cover risks
```{r}
res_long %>%
  mutate(label = paste0(outcome, " (ranked cor ", scorr,")")) %>%
  
  ggplot(aes(x=cover, y=MPR, colour=diff_cat, shape=diff_cat)) + 
  geom_point(alpha=.5) + 
  geom_abline(intercept=0, slope=1, colour='black') + 
  geom_abline(intercept=-.1, slope=1, colour='grey50', linetype=2) + 
  geom_abline(intercept=.1, slope=1, colour='grey50', linetype=2) + 
  
  facet_grid(.~label) +
  coord_fixed() +
  
  scale_colour_colorblind(name='') +
  scale_shape_discrete(name='') + 
  scale_y_continuous(lim=c(0,1), "19andMe") +
  scale_x_continuous(lim=c(0,1), "COVER") +
  
  theme_bw(base_size=10) + 
  theme(panel.grid.major = element_blank(),  
        panel.grid.minor = element_blank(),  
        legend.position='bottom')

ggsave(file.path("..", "figures", "figure-cover-01.jpeg"), height=3, width=6)
```

### Plot 19andMe versus cover risks - no conditions
```{r}
filter(res_long, !has_condition) %>%
  
  ggplot(aes(x=cover, y=MPR, colour=diff_cat, shape=diff_cat)) + 
  geom_point(alpha=.5) + 
  geom_abline(intercept=0, slope=1, colour='black') + 
  geom_abline(intercept=-.1, slope=1, colour='grey50', linetype=2) + 
  geom_abline(intercept=.1, slope=1, colour='grey50', linetype=2) + 
  
  geom_point(data = res_long, alpha=0) + # ghost points for legend (boo!)
  
  facet_grid(.~outcome) +
  coord_fixed() +
  
  scale_colour_colorblind(name='') +
  scale_shape_discrete(name='') + 
  scale_y_continuous(lim=c(0,.3), "19andMe", labels = scales::percent) +
  scale_x_continuous(lim=c(0,.3), "COVER", labels = scales::percent) +
  
  theme_bw() + 
  theme(panel.grid.major = element_blank(),  
        panel.grid.minor = element_blank(),  
        legend.position='bottom')

#ggsave(file.path("..", "figures", "figure-cover-02.pdf"), height=5, width=8)
```

## Base rates by age and sex
```{r}
filter(res_long, !has_condition) %>%
  mutate(sex = str_to_sentence(sex)) %>%
  gather("app","score", c(MPR, cover)) %>%
  filter(app!="MPR" | age%%10==0 & age<=80) %>%
  
  ggplot(aes(x=age, y=score, color=app, shape=sex)) +
  geom_point(alpha=.5) +
  geom_line(alpha=.5) +
  
  facet_grid(.~outcome) +
  scale_y_continuous("Probability of adverse outcome") +
  scale_color_discrete("",labels = c("COVER","19andMe")) +
  scale_shape_discrete("") +
  
  theme_bw(base_size = 10) +
  theme(panel.grid.major = element_blank(),  
        panel.grid.minor = element_blank(),  
        legend.position='bottom')

ggsave(file.path("..", "figures", "figure-cover-02.jpeg"), height=3, width=6)
```

## Number of conditions
```{r}
filter(res_long, diff_cat!="19andMe 10% lower") %>%
  mutate(age_cat = ifelse(age>=60, "Age 60+", "Under age 60")) %>%
  ggplot(aes(x=conditions_num, fill = diff_cat)) +
  geom_bar(position = position_dodge2(preserve="single")) +
  facet_grid(age_cat~outcome) +
    
  scale_fill_manual(name='', values=c('grey30', 'grey70')) + 
  xlab("Number of conditions") +
  scale_y_continuous("", labels = scales::comma) +
  
  theme_bw(base_size=10) +
  theme(panel.grid.major = element_blank(),  
        panel.grid.minor = element_blank(),  
        legend.position='bottom')

ggsave(file.path("..", "figures", "figure-cover-03.jpeg"), height=3, width=6)
```

## Specific conditions
```{r}
filter(res_long, diff_cat!="19andMe 10% lower") %>%
  pivot_longer(COND_VARS_ALL, names_to = "condition",values_to = "present")  %>% 
  filter(condition!="is_other") %>%
  mutate(in_cover = ifelse(condition %in% COND_VARS_COVER, "In COVER", "Not in COVER")) %>%
  group_by(outcome, diff_cat, condition, in_cover) %>%
  summarise(p = mean(present)) %>%
  
  ggplot(aes(x=condition, y=p, fill=diff_cat)) +
  geom_col(position = "dodge") +
  facet_grid(outcome~in_cover, scales = "free_x") +
  theme_bw() + 
  theme(legend.position = "bottom")
```



## Statistical tests

### Wilcoxen test for number of preexisting conditions
```{r, results="asis"}
wilcox_test_fmt <- function(x,y, bonf_n = 1){
  #print(wilcox.test(unlist(x),unlist(y)))
  format.pval(wilcox.test(unlist(x),unlist(y))$p.value*bonf_n, nsmall=3, ndigits=3, eps=1e-3)
}

# Wilcoxon test to compare mean number of pre-existing conditions 
res_long %>%
  mutate(age_cat = ifelse(age>=60, "60+", "Under 60")) %>%
  select(outcome, age_cat, diff_cat, conditions_num) %>%
  mutate(diff_cat = gsub(" |\\%|\\d","",diff_cat)) %>%
  nest(conditions_num = conditions_num) %>%
  pivot_wider(names_from = diff_cat, values_from = conditions_num) %>%
  mutate(pval = map2(within, andMehigher, wilcox_test_fmt)) %>%
  
  select(outcome, age_cat, pval) %>%
  unnest(pval) %>%
  pandoc.table()
```

## Specific conditions
```{r}
prop_test_fmt <- function(df, bonf_n = 1){
  #print(prop.test(x=df$x, n=df$n))
  format.pval(prop.test(x=df$x, n=df$n)$p.value*bonf_n, nsmall=3, ndigits=3, eps=1e-3)
}

res_long %>%
  mutate(age_cat = ifelse(age>=60, "60+", "Under 60")) %>%
  select(outcome, age_cat, diff_cat, COND_VARS_ALL) %>% 
  filter(diff_cat!="19andMe 10% lower") %>%
  pivot_longer(COND_VARS_ALL, names_to = "condition", values_to = "present") %>%
  filter(condition!="is_other") %>%
  group_by(outcome, age_cat, diff_cat, condition) %>%
  summarise(x=sum(present), n=n()) %>%
  nest(props = c(diff_cat, x, n)) %>%
  mutate(pval = map(props, prop_test_fmt, bonf_n = 8)) %>%
  
  select(outcome, condition, pval) %>%
  unnest(pval) %>%
  pivot_wider(names_from = outcome, values_from = pval) %>%
  pandoc.table(split.table=Inf)
```
