---
title: "missing_api"
author: "Emma Pendl-Robinson"
date: "10/21/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
source('api_helper_functions.R')
```

Example of POST Request

```{r, eval=FALSE}

request_url <- "api.covid19.mathematica.org/score"

request_body  <- list(
  "zip" = "64034",
  "age"= 29,
  "sex" = "other", 
  "symptoms" = as.list(list('is_other')),
  "nppl" = 1,
  "is_roommate"= "True",
  "nppl2"= 10,
  "hand"= "True",
  "ppe"= "True",
  "conditions" =  as.list(list('is_cvd'))
  )

jsonlite::toJSON(request_body , auto_unbox = TRUE)


r <- POST(request_url, add_headers("x-api-key" = Sys.getenv("X_API_KEY")), body = request_body, encode = "json")
content(r)
```

# who were the people who I could not get a score for

```{r}
# import nexoid data
nexoid_for_api_input <- readRDS("N:/Transfer/JStarling/19andme/nexoid_data_forapi.RData")
# get a list of all the person from the nexiod data
input_person_names <- names(nexoid_for_api_input) 
```


```{r}
# define column names
csv_cols <- cols(
  person = col_character(),
  fips = col_character(),
  name = col_character(),
  n_case_today = col_double(),
  latest_day = col_character(),
  moving_casecount = col_double(),
  underreport_factor = col_double(),
  est_unreported_sick = col_double(),
  exposure_risk = col_double(),
  sympt_covid_risk = col_double(),
  hosp_risk = col_double(),
  icu_risk = col_double(),
  death_risk = col_double(),
  vulnerability_risk = col_double(),
  score = col_double()
)

#read in the results from N:\Transfer\JStarling\19andme\api_output
file_paths <- list.files(path = "N:/Transfer/JStarling/19andme/api_output/", pattern = "csv_part_", recursive = TRUE, full.names = TRUE)
df <- map_dfr(file_paths, ~ read_csv(.x, col_types = csv_cols))
  
```

```{r}
# get a list of all the persons from the df
df_person_names <- df$person 

# find people in input nexoid but not df
missing_people <- input_person_names[!input_person_names %in%df_person_names]
length(missing_people)

#filter nexoid data to missing people
missing_ls <- nexoid_for_api_input[!input_person_names %in%df_person_names]

```

make dataframe with the missing people's information


```{r}
vec2str <- function(x){
  # convert vector of conditions to a string

  # x : list of person attributes
  x$conditions <- toString(x[['conditions']])
  return(x)
}

missing_ls_vec_conditions <- lapply(missing_ls, vec2str)

#making a dataframe with the data to make it more human readable
missing_df <- list2_df(missing_ls_vec_conditions)

missing_df
```

running calc risk for the missing people and figure out where/what the erorrs are

```{r}
calculateRisk_miss <- function(input) {
  #makes API Calls
  request_body <- list(
    'zip' = input[['zip']],
    'age' = as.numeric(input[['age']]),
    'sex' = gsub("is_", "", input[['sex']]),
    'symptoms' = na_input(input[['symptoms']]),
    'nppl' = as.numeric(input[['nppl']]),
    'is_roommate' = bol2char(input[['is_roommate']]),
    'nppl2' = as.numeric(input[['nppl2']]),
    'hand' = bol2char(input[['hand']]),
    'ppe' = bol2char(input[['ppe']]),
    'conditions' = na_input(input[['conditions']])
  )
  #print(request_body)
  #print(jsonlite::toJSON(request_body , auto_unbox = TRUE))
  resp <- POST(request_url, add_headers("x-api-key" = Sys.getenv("X_API_KEY")), body = request_body, encode = "json")
  api_return <- content(resp)
  
  return(api_return)
}

# calcualte risk for all the missing people with lapply

plan(multiprocess) ## => parallelize on your local computer
miss_out <- future_lapply(missing_ls, calculateRisk_miss)
```

filter to outputs with error messages

```{r}
error_message <- miss_out[lapply(miss_out, length) == 1]

error_message_df <- list2_df(error_message)
```

More than one fips is returned that does not equal Jackson County, MO
logic:
- if one of the api results matches the nexoid fip, select this one
- if none of the api results matches the nexoid fip and there is only one api result fips, select the one api result
- if none of the api results matches the nexoid fip and more than one of the api return county name is Jackson County, MO then return the first api result
- if none of the conditions are meet, return NULL
```{r}
# flatten to make a df
# https://stackoverflow.com/questions/48542874/flatten-nested-list-of-lists-with-variable-numbers-of-elements-to-a-data-frame

two_results <- lapply(miss_out[lapply(miss_out, length) == 2],'[[', 'results')

two_results_nest_df  <- map(map_if(two_results,~class(.x)=="matrix",list),~map(.x,as.data.frame))

two_results_df <- map_dfr(two_results_nest_df,~map_dfr(.x,identity),identity,.id="persons")

```


Save out error message df and the two results df

add the missing_results_out two_results_df to the df of all the nexoid results and save
```{r}
write_csv(error_message_df, "N:/Transfer/JStarling/19andme/api_output/error_message.csv")
write_csv(two_results_df, "N:/Transfer/JStarling/19andme/api_output/two_results_no_match.csv")

```

```{r}
all_df <- bind_rows(df, two_results_df)

write_csv(all_df, "N:/Transfer/JStarling/19andme/api_output/all_api_output.csv")

```


